---
title: "Finlay-Wilkinson Regression"
author: "neyhartj"
date: "2023-01-10"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


## Introduction

Finlay-Wilkinson regression is a popular statistic approach to analyze linear genotype-environment interactions. The traditional regression approach used ordinary least squares, but mixed-model versions can handle situations like heterogeneity of genotypic deviations from regression or missing data.



## FW Regression with Environmental Covariates

This is an R code version of the Example 1 SAS code provided by Piepho.

The format will include the SAS code and the R equivalent


This example looks at a lettuce dataset. The phenotypic data contains

```{r load.lettuce}
load("data/vanEeuwijk1992Lettuce.RData")
lettuce_pheno <- pheno %>%
  mutate(environment = fct_inseq(environment),
         genotype = as.factor(genotype)) %>%
  arrange(across())

# Convert to wide format
lettuce_pheno_wide <- lettuce_pheno %>%
  mutate(genotype = factor(genotype, levels = c("Pa", "DM", "Pi", "GT", "RW", "Wi", "Tr", "Ls"))) %>%
  spread(genotype, nitrate)

lettuce_ec <- ec %>%
  mutate(environment = fct_inseq(environment)) %>%
  # Rename to be consistent with Piepho
  rename_all(~str_replace(., "ec", "x"))

# Standardize the ec data
lettuce_ec_std <- lettuce_ec %>%
  mutate_at(vars(-environment), ~as.numeric(scale(.)))

# Merge the data
lettuce <- left_join(lettuce_pheno, lettuce_ec_std) %>%
  rename(gen = genotype, env = environment)

# Ignore singularities
asreml.options(ai.sing = TRUE, workspace = 1e9)

```


#### Baseline model

SAS code:

```{sas}
/*BASELINE*/
proc mixed data=g covtest method=REML;
class gen env;
model yield=gen/noint solution; # noint omits the intercept
random env;
repeated env/ sub=gen type=ar(1);
run;

```

R version:

```{r}
fit1 <- asreml(fixed = nitrate ~ -1 + gen, # No intercept
               random = ~ env, 
               residual = ~ ar1(env):gen,
               data = lettuce)

fit1_summ <- summary(fit1)

# Print variance components
fit1_summ$varcomp


```

These match the variance components presented in Table 5 for the "Without covariate" section.


#### FA model


SAS code:

```{sas}
/*FA*/

proc mixed data=a lognote;
class gen env;
model yield=gen env;
run;

proc mixed data=a lognote;
class gen env;
model yield=gen env;
repeated env/sub=gen type=ar(1);
run;

```

R version:

```{r}
fit2a <- asreml(fixed = nitrate ~ gen + env, 
                data = lettuce)
fit2a_summ <- summary(fit2a)

fit2b <- asreml(fixed = nitrate ~ gen + env, 
                residual = ~ ar1(env):gen,
                data = lettuce)
fit2b_summ <- summary(fit2b)


```


#### FA0(1) model

##### Different version of fitting FA models

We can fit a FA model using the `fa()` special function in asreml:

```{r}

# First, fit a regular FA model using the fa argument
fit_fa1 <- asreml(fixed = nitrate ~ gen + env,
                  random = ~ fa(env, k = 2):gen,
                  residual = ~ ar1(env):gen,
                  data = lettuce)

# # Second, fit a fa model using the rr special and the diagonal special
# fit_fa2 <- asreml(fixed = nitrate ~ gen + env, 
#                   random = ~ rr(env, k = 2):gen + diag(env):gen,
#                   residual = ~ ar1(env):gen,
#                   data = lettuce)


```



##### Fit the models



SAS code:

```{sas}
     /*FA0(1)*/

proc mixed data=a lognote covtest method=ML;
ods output covparms=cp0;
class gen env;
model yield=gen env;
random x1-x8/sub=gen type=FA0(1);
parms / pdata=cp;
parms (.1)(.1)(.1)(.1)(.1)(.1)(.1)(.1)
      (.1)(.1);
repeated env/sub=gen type=ar(1);
run;

```

R code:

```{r}
# Assume EC is a pxq matrix of known environmental covariates
EC_df <- lettuce %>%
  select(env, starts_with("x")) %>%
  distinct() %>%
  as.data.frame()
EC <- as.matrix(column_to_rownames(EC_df, "env"))
# 
# # Fake A matrix
# A <- diag(nlevels(lettuce$gen))
# dimnames(A) <- list(levels(lettuce$gen), levels(lettuce$gen))
# 
# fit_fa2 <- asreml(fixed = nitrate ~ gen + env, 
#                   random = ~ str( ~ mbf(EC):vm(gen, A), vmodel = ~ rr(7, 1):vm(gen, A)),
#                   residual = ~ ar1(env):gen,
#                   mbf = list(EC = list(key = c("env", "env"), cov = "EC_df")),
#                   data = lettuce)


```

R code:

```{r}

lettuce1 <- lettuce %>% 
  mutate(x0 = 0,
         x01 = x0)

# Random effect formulation
random_form <- reformulate(paste0(sort(grep(pattern = "^x", x = names(lettuce1), value = TRUE)), ":gen"))

# fit_fa2 <- asreml(fixed = nitrate ~ -1 + gen + env, 
#                   random = ~ str(form = ~x0:gen + x01:gen + x1:gen + x2:gen + x3:gen + x4:gen + x5:gen + x6:gen + x7:gen + x8:gen, vmodel = ~ rr(8, k = 2):id(8)),
#                   residual = ~ ar1(env):gen,
#                   data = lettuce1)


```



proc mixed data=a lognote covtest method=REML;
ods output covparms=cp;
class gen env;
model yield=gen env;
random x1-x8/sub=gen type=FA0(1);
parms / pdata=cp0;
repeated env/sub=gen type=ar(1);
run;

data cp2;
set cp;
output;
if _N_<9 and _N_>1 then do;
  estimate=0.1; output;
end;
run;

proc mixed data=a lognote;
ods output covparms=cp3;
class gen env;
model yield=gen env;
random x1-x8/sub=gen type=FA0(2);
parms / pdata=cp2;
repeated env/sub=gen type=ar(1);
run;

data cp;
set cp;
if _N_<9;
array slope slope1-slope8;
do i=1 to 8;
  slope[i]=0;
end;
slope[_N_]=estimate;
run;

proc means noprint data=cp;
var slope1-slope8;
output out=slopes sum=;
run;

data g;
set a;
if _N_=1 then set slopes;
array slope slope1-slope8;
array x x1-x8;
z=0;
do i=1 to 8;
  z=z+slope[i]*x[i];
end;
run;

proc mixed data=g method=ML;
class gen env;
model yield=gen z gen*z /ddfm=KR;
random env;
repeated env/ sub=gen type=ar(1);
run;

proc mixed data=g method=REML covtest;
class gen env;
model yield=gen gen*z/noint solution;
random env;
repeated env/ sub=gen type=ar(1);
run;


R version:

```{r}


fit3a <- asreml(fixed = nitrate ~ gen + env, 
                data = lettuce)


fit2a_summ <- summary(fit2a)

fit2b <- asreml(fixed = nitrate ~ gen + env, 
                residual = ~ ar1(env):gen,
                data = lettuce)
fit2b_summ <- summary(fit2b)


```


proc mixed data=a lognote covtest method=ML;
ods output covparms=cp0;
class gen env;
model yield=gen env;
random x1-x8/sub=gen type=FA0(1);
parms / pdata=cp;
parms (.1)(.1)(.1)(.1)(.1)(.1)(.1)(.1)
      (.1)(.1);
repeated env/sub=gen type=ar(1);
run;

proc mixed data=a lognote covtest method=REML;
ods output covparms=cp;
class gen env;
model yield=gen env;
random x1-x8/sub=gen type=FA0(1);
parms / pdata=cp0;
repeated env/sub=gen type=ar(1);
run;



#### SVD

SAS code:

```{sas}
/*SVD*/

data svd;
set svd;
keep gen env pred;
run;

proc iml;
use svd; 
read all var _num_ into pred;
pred=pred[,2];
close svd;
use svd0; 
read all var _num_ into int;
int=int[1:8,1];
print int;
close svd0;
p=shape(pred,18,8); p=t(p);
p_centered=p-int*j(1,18,1);
print pred p p_centered;
call svd(u,q,v,p_centered);
print u q v;
a=u*diag(q)*t(v);
print a;
create z_svd from v;
append from v;
close z_svd;
run;
quit;

data z_svd;
set z_svd;
z1=col1;
z2=col2;
z3=col3;
env=_N_;
run;

       /*check*/
# data w;
# merge e z_svd;
# by env;
# run;
# 
# proc reg;
# model col1=x1-x8;
# run;

data v;
merge a z_svd;
by env;
run;

proc mixed data=v method=REML covtest;
class gen env;
model yield=gen gen*z1/noint solution;
random env;
repeated env/sub=gen type=ar(1);
run;

proc mixed data=v method=ML;
class gen env;
model yield=gen gen*z1;
random env;
repeated env/sub=gen type=ar(1);
run;

proc mixed data=v method=REML;
class gen env;
model yield=gen gen*z1 gen*z2;
random env;
repeated env/sub=gen type=ar(1);
run;

proc mixed data=v method=ML;
class gen env;
model yield=gen gen*z1 gen*z2;
random env;
repeated env/sub=gen type=ar(1);
run;



```

R code:

```{r}

# SVD of environmental covariates
ec_mat <- lettuce_ec_std %>%
  column_to_rownames("environment") %>%
  as.matrix()

ec_svd <- svd(x = t(ec_mat))
z_svd <- `dimnames<-`(ec_svd$v, list(row.names(ec_mat), paste0("z", seq_len(ncol(ec_mat)))))

lettuce1 <- left_join(select(lettuce, -starts_with("x")), rownames_to_column(as.data.frame(z_svd), "env")) %>%
  mutate(env = factor(env, levels = levels(lettuce$env)))


# proc mixed data=v method=REML covtest;
# class gen env;
# model yield=gen gen*z1/noint solution;
# random env;
# repeated env/sub=gen type=ar(1);
# run;

fit_svd1 <- asreml(fixed = nitrate ~ gen + gen:z1, 
                   random = ~ env,
                   residual = ~ ar1(env):gen,
                   data = lettuce1)

fit_svd1_summ <- summary(fit_svd1)

fit_svd1_summ$varcomp


fit_svd2 <- asreml(fixed = nitrate ~ gen + gen:z1 + gen:z2, 
                   random = ~ env,
                   residual = ~ ar1(env):gen,
                   data = lettuce1)

fit_svd2_summ <- summary(fit_svd2)

fit_svd2_summ$varcomp
```




#### Factorial regression

SAS code:

```{sas}
/*FACTORIAL REGRESSION*/

proc mixed data=a covtest method=REML;
ods output solutionF=svd0;
class gen env;
model yield=gen x1-x8 gen*x1 gen*x2 gen*x3 gen*x4 gen*x5 gen*x6 gen*x7 gen*x8/noint solution ddfm=KR outpm=SVD;
random env;
repeated env/ sub=gen type=ar(1);
parms (.1)(.1)(.1)/lowerb=0,0,0;
run;

proc mixed data=a covtest method=ML;
ods output solutionF=svd0;
class gen env;
model yield=gen x1-x8 gen*x1 gen*x2 gen*x3 gen*x4 gen*x5 gen*x6 gen*x7 gen*x8/noint solution ddfm=KR outpm=SVD;
random env;
repeated env/ sub=gen type=ar(1);
parms (.1)(.1)(.1)/lowerb=0,0,0;
run;


```

R code:

```{r}

# Create the fixed formula
fixed_form <- reformulate(termlabels = c("gen", names(lettuce_ec)[-1], paste0("gen:", names(lettuce_ec)[-1])),
                          response = "nitrate")

fit_fr1 <- asreml(fixed = fixed_form, 
                  random = ~ env,
                  residual = ~ ar1(env):gen,
                  data = lettuce)

fit_fr1_summ <- summary(fit_fr1)

fit_fr1_summ$varcomp

```

These match the variance components presented in Table 5 for the "with covariates x1-x8" FR section.




#### PLS

##### Run the PLS on the environmental covariates

SAS code:


```{sas}
/*PLS*/

proc standard data=e out=f2 mean=0 std=1;
var x1-x8;
run;

data h;
merge f2 wide;
by env;
run;

proc pls data=h method=pls noscale nocenter;
model gen1-gen8=x1-x8;
output out=z XSCORE=z;
run;

```


R version:

```{r}
library(pls)

# lettuce_ec_std is the standardized EC data frame
# 
# merge with the lettuce pheno wide
lettuce_wide1 <- left_join(lettuce_pheno_wide, lettuce_ec_std) %>%
  rename(env = environment)

# Separate into matrices
Y <- as.matrix(column_to_rownames(lettuce_pheno_wide, "environment"))
X <- as.matrix(column_to_rownames(lettuce_ec_std, "environment"))

# Run PLS
z_pls <- plsr(Y ~ X, scale = FALSE, center = FALSE)

# Get the scores
z_pls_scores <- scores(z_pls)[,] %>% 
  as.data.frame() %>%
  rename_all(~paste0("z", seq_along(.))) %>%
  rownames_to_column("env") %>%
  mutate(env = as.factor(env))
  


```


SAS code:

```{sas}

data k;
merge z b;
by env;
run;

proc mixed data=k;
class gen env;
model yield=gen z1 gen*z1 /ddfm=KR;
random env;
repeated env/ sub=gen type=ar(1);
run;

proc mixed data=k covtest method=REML;
class gen env;
model yield=gen gen*z1/noint solution;
random env;
repeated env/ sub=gen type=ar(1);
run;

proc mixed data=k covtest method=ML;
class gen env;
model yield=gen gen*z1/noint;
random env;
repeated env/ sub=gen type=ar(1);
run;

proc mixed data=k covtest method=REML;
class gen env;
model yield=gen gen*z1 gen*z2/noint solution;
random env;
repeated env/ sub=gen type=ar(1);
run;

proc mixed data=k covtest method=ML;
class gen env;
model yield=gen gen*z1 gen*z2/noint;
random env;
repeated env/ sub=gen type=ar(1);
run;

```

R code:

```{r}
# data k;
# merge z b;
# by env;
# run;
lettuce1 <- left_join(lettuce, z_pls_scores)


# proc mixed data=k covtest method=REML;
# class gen env;
# model yield=gen gen*z1/noint solution;
# random env;
# repeated env/ sub=gen type=ar(1);
# run;
fit_pls1 <- asreml(fixed = nitrate ~ -1 + gen + gen:z1, 
                   random = ~ env,
                   residual = ~ ar1(env):gen,
                   data = lettuce1)


fit_pls2 <- asreml(fixed = nitrate ~ gen + gen:z1 + gen:z2, 
                   random = ~ env,
                   residual = ~ ar1(env):gen,
                   data = lettuce1)



fit2 <- asreml(fixed = nitrate ~ -1 + gen + gen:z1, # No intercept
               random = ~ env + ar1(env):gen, 
               residual = ~ units,
               data = lettuce1)


```

These match the variance components presented in Table 5 for the "With covariate z1" under PLS











